\chapter{Integer Transition Systems}
\chapterauthor{Florian Frohn}

\begin{abstract}
  This chapter describes the syntax and semantics for \emph{Integer Transition Systems}.
  %
  Such systems are \emph{Logically Constrained Term Rewrite Systems} (LCTRSs) where the underlying SMT theory is fixed to the theory of integers, and where function symbols must not occur anywhere but at the root position.
  %
  Due to the latter restriction, we only consider top-level rewriting.
  %
  Intuitively, such programs correspond to \texttt{while}-programs (or, equivalently, tail-recursive programs) operating on integers.
\end{abstract}

\section{Syntax}

\subsection{Grammar}
\

\begin{tabular}{lll}
its & ::= & \texttt{(format LCTRS) (theory Ints)} fun+ entrypoint rule+ \\
fun & ::= & \texttt{(fun} IDENTIFIER \texttt{(->} int+\texttt{))} \\
int & ::= & \texttt{Int} \\
entrypoint & ::= & \texttt{(entrypoint} IDENTIFIER\texttt{)} \\
lhs & ::= & IDENTIFIER $|$ \texttt{(}IDENTIFIER IDENTIFIER+\texttt{)} \\
rhs & ::= & IDENTIFIER $|$ \texttt{(}IDENTIFIER int-term+\texttt{)} \\
guard & ::= & \texttt{:guard} formula \\
var & ::= & \texttt{:var (}IDENTIFIER type\texttt{)} \\
rule & ::= & \texttt{(rule} lhs rhs guard? var*\texttt{)} \\
\end{tabular}

\bigskip
Here, ``int-term'' refers to a term of type \texttt{Int} from the underlying SMT-theory \texttt{Ints}, and ``formula'' refers to a term of type \texttt{Bool}.
%
The \texttt{:var} statements are required if the types of variables cannot be inferred (e.g., if the only occurrences of \texttt{x,y} are in the literal \texttt{(= x y)}).

\subsection{Restrictions}

The IDENTIFIER in ``entrypoint'' must be one of the previously declared functions.

\subsection{Remarks}

We allow arbitrary terms and formulas from the \texttt{Ints} theory.
%
In particular, this means that we allow:
\begin{itemize}
\item non-linear arithmetic
\item exponentiation
\item quantifiers
\item \texttt{ite} (if-then-else) expressions
\item \texttt{let} expressions
\item variables of type \texttt{Bool}
\item all usual Boolean connectives (conjunction, disjunction, negation, implication, $\ldots$)
\end{itemize}

We do not require left-linearity, so, e.g.,
\begin{center}
  \texttt{(rule (f x x) (g x))}
\end{center}
is a valid rule.

\section{Semantics}

Let $\F$ be the set of all declared function symbols (i.e., we have $\mathtt{f} \in \F$ if the input file contains a line \texttt{(fun f \ldots)}).
%
In the sequel, we write $f(t_1,\ldots,t_n)$ instead of $\mathtt{(f\ t_1\ \ldots\ t_n)}$ for the sake of readability.
%
Moreover, we always assume that terms respect the arity of the respective function symbols (i.e., if we write $f(t_1,\ldots,t_n)$, then we implicitly assume that $f$ has been declared as $\mathtt{(fun\ f\ ({-}{>}\ \underbrace{\mathtt{Int\ \ldots\ Int}}_{n \times}\ Int))}$.
%
For a rule with left-hand side $\ell$, right-hand side $r$, and guard $\varphi$, we write $\ell \to r\ \guard{\varphi}$.
%
$\Rules$ denotes the set of all rules that are defined in the input file and $\V$ contains all variables that occur in the input file.
%
The unique entrypoint (as defined by the line \texttt{(entrypoint ...)}) is denoted $\mathit{entry}$.
%
A \emph{configuration} is a term $f(c_1,\ldots,c_n)$ where $f \in \F$ and $c_1,\ldots,c_n \in \Z$.
%
A \emph{valuation} is a mapping $\sigma: \V \to \Z$.
%
$\Rules$ induces a rewrite relation on configurations as follows:
%
We have
\[
  f(c_1,\ldots,c_n) \to_\Rules g(d_1,\ldots,d_m)
\]
if there is a rule $f(x_1\ldots,x_n) \to g(t_1,\ldots,t_m)\ \guard{\varphi} \in \Rules$ and a valuation $\sigma$ such that $\sigma(x_i) = c_i$ for all $i \in \{1, \ldots, n\}$, $t_i\sigma \equiv_{\mathtt{Ints}} d_i$ for all $i \in \{1, \ldots, m\}$, and $\sigma \models_{\mathtt{Ints}} \varphi$.
%
Here, $\sigma \models_{\mathtt{Ints}} \varphi$ means that there is an extension $\theta \in \mathtt{Ints}$ of $\sigma$ such that $\theta \models \varphi$ (note that SMT theories are defined as classes of structures), and $t_i\sigma \equiv_{\mathtt{Ints}} d_i$ means that we have $\theta \models t_i\sigma = d_i$ for all $\theta \in \mathtt{Ints}$.
%
A \emph{run} is a (possibly infinite) sequence
\[
  \mathit{entry}(c_1,\ldots,c_n) \to_\Rules s_1 \to_\Rules s_2 \to_\Rules \ldots
\]
An integer transition system terminates if all runs are finite.